Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-16 demystify.tex
updates

% Prelim work does not, and in fact should not, be extensive. You do not need a paper or a complete implementation.  You don't need to show that your idea is better than others or even that it will work.  You just need to show that you have tried it and it has some promise.  This can be as simple as a small experiment done by hand on some small examples,  e.g., if you come up with a new abstraction in PL for proving correctness, then you can just manually apply it to a few small programs, preferrably ones that existing works have trouble with.  This is enough to show that you have thought about the problem and have some initial results.

% To make prelim concrete (and thus more convincing), you might want to use specific names and numbers, e.g.,  \emph{``I tried my prototype on the MNIST dataset, achieving an accuracy of 90\%''}, or \emph{``I used Lean to formalize and prove a specific case of the Fermat's Last Theorem''}.  In addition, negative results are results and can be very convincing.  So don't be afraid to say \emph{``I tried my idea on X but it failed because of Y, and I plan to address this by Z''}.

\paragraph{Research} is the emphasis.  Many made a mistake of focusing on teaching and mentoring and not on research. This is especially true for students who have a lot of teaching and mentoring experience, e.g., from a teaching-focused university or a teaching assistant.  While teaching and mentoring are important, the GRFP is a \emph{research} fellowship which aims to train researchers. So you will need to demonstrate your passion for research and your potential to be a good researcher.

\paragraph{Overexaggeration Lors} Many students have LoRs that overrate them, e.g., \emph{``the best student I have ever had''}, \emph{``better than most PhD students''}, etc.  This is not only not convincing but can also hurt your application.  Reviewers are often professors themselves and know that such statements are often not true.  Moreover, such statements can bias the evaluation, e.g., if a reviewer sees that a student is from a top school and has a LoR from a top professor, they might give the student a higher score than they deserve.  So you should ask your LoR writers to be honest and specific, e.g., \emph{``the student did well in my class and I think they have the potential to be a good researcher''}.

>>>>

\paragraph{Related Work} Many research plans motivate the problem well but do not discuss limitations of existing work, making it questionable if the problem is important or if you are aware of the challenges.  Thus, you should do a thorough literature review and discuss what people have done and their limitations.  Your research plan should then talk about how you plan to address these limitations, i.e., fill the gap in the literature.

\paragraph{Do NOT BS} Many research plans include technical details or proposed work that are vauge, unrealistic, or even impossible.  This is similar to a student taking an exam and write a lot of nonsense to fill the page and hope for partial credits.  As mentioned, reviewers are often chosen based on their expertise in your field and can detect B.S. Again, do your homework and ask someone who knows the field, e.g., your advisor or LoR writer, to review your research plan.

\paragraph{Overexaggeration Lors} Many GRFP applicants are from top schools with well-known professors that are very good at writing. The issue is that they often use flowery language and overrate their students. Reviewers are often professors themselves and can detect B.S. So you should ask your LoR writers to be honest, tone down the language (similar to overclaiming research results when writing papers), and be specific with concrete examples to demonstrate why you're ``the best''.
